Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2025-01-21 LICS/sections/appendix.tex
wip

\section{$\catname{SLat}$}
\label{sec:appendix:1}

@@

\section{Proofs for Section \ref{sec:e-hypergraphs}: E-hypergraphs}
\label{sec:appendix:2}

In this section we will define the conditions under which the pushout in $\catname{EHyp}(\Sigma)$ exists and we will explicitly construct such a pushout.
We will also show the uniqueness of a boundary pushout complement.
These condition will play a crucial role in showing that for each enrichment equation in $\textbf{PROP}^{+}(\Sigma)$ there is a sequence of rewrites in $\WellTypedMdaEcospans$.

@@

    Let's show the direction from left to right.
    Note that $f_{ext}$ and $f_{ext}'$ have inverses since they are injective and surjective (the injection is per the definition and surjection is because domain and codomain are equal).
    Then we need $f_{ext};\beta = f_{ext}'$ and we define $\beta = f_{ext}^{-1};f_{ext}'$.
    Similarly $\beta^{-1} = f_{ext}'^{-1};f_{ext}$. 
    We can check that $\beta^{-1};\beta = f_{ext}'^{-1};f_{ext};f_{ext}^{-1};f_{ext}' = f_{ext}'^{-1};id;f_{ext}' = f_{ext}'^{-1};f_{ext}' = id$.
    Then we have that $f_{ext};\beta;f_{int}' = f_{ext};f_{ext}^{-1};f_{ext}';f_{int}' = f_{ext}';f_{int}' = f' = f;\alpha = f_{ext};f_{int};\alpha$.
    Since $f_{ext}$ is surjective, we have $\beta;f_{int}' = f_{int};\alpha$.
    Analogously for $\beta^{-1}$ and the right-hand sides of the diagrams.
\end{proof}

@@

\section{Additional figures for Section~\ref{sec:e-hypergraphs}: E-hypergraphs}
\begin{example}
    Consider cospans from Figure~\ref{fig:appendix:non-isomorphic-cospans}(a) where the $f_{int}, f_{ext}$ (respectively, $g_{int},g_{ext}$) maps are shown with red arrows.
    The left cospan represents a symmetry placed inside a hierarchical edge while the right one depicts a tensor product of two identities put inside the hierarchical edge.
    The colours of the vertices show the induced partition of the interfaces.
    We will try to build an isomorphism step-by-step.
    To make the diagram commute, clearly, it must be the case that $u_0 \mapsto u_0$ and $u_1 \mapsto u_1$.
    Then, because the isomorphism should preserve the order within the subset of green verices, it must be the case that $u_2 \mapsto u_2$ and $u_3 \mapsto u_3$.
    Similarly for the output interfaces.
    To further make the diagram commute, we need to map $v_2 \mapsto v_3$ and $v_3 \mapsto v_2$.
    This, however, contradicts the requirement that $g_{int};\gamma(w_2) = g_{int}(w_2)$.
    Hence, the two cospans are not isomorphic.
\end{example}

>>>>

\section{$\catname{SLat}$}
\label{sec:appendix:slat}

@@

\section{Proofs for Section \ref{sec:e-hypergraphs}: E-hypergraphs}
\label{sec:appendix:pushout}

In this section we will define the conditions under which the pushout in $\catname{EHyp}(\Sigma)$ exists, and we will explicitly construct such a pushout.
We will also show the uniqueness of a boundary pushout complement.
These condition will play a crucial role in showing that for each enrichment equation in $\textbf{PROP}^{+}(\Sigma)$ there is a sequence of rewrites in $\WellTypedMdaEcospans$.

@@

    Let's show the direction from left to right.
    Note that $f_{ext}$ and $f_{ext}'$ have inverses since they are injective and surjective (the injection is per the definition and surjection is because domain and codomain are equal).
    Then we need $f_{ext};\beta = f_{ext}'$ and we define $\beta = f_{ext}^{-1};f_{ext}'$.
    Similarly, $\beta^{-1} = f_{ext}'^{-1};f_{ext}$. 
    We can check that $\beta^{-1};\beta = f_{ext}'^{-1};f_{ext};f_{ext}^{-1};f_{ext}' = f_{ext}'^{-1};id;f_{ext}' = f_{ext}'^{-1};f_{ext}' = id$.
    Then we have that $f_{ext};\beta;f_{int}' = f_{ext};f_{ext}^{-1};f_{ext}';f_{int}' = f_{ext}';f_{int}' = f' = f;\alpha = f_{ext};f_{int};\alpha$.
    Since $f_{ext}$ is surjective, we have $\beta;f_{int}' = f_{int};\alpha$.
    Analogously for $\beta^{-1}$ and the right-hand sides of the diagrams.
\end{proof}

@@

\section{Additional figures for Section~\ref{sec:e-hypergraphs}: E-hypergraphs}
\label{sec:appendix:iso}
\begin{example}
    Consider cospans from Figure~\ref{fig:appendix:non-isomorphic-cospans}(a) where the $f_{int}, f_{ext}$ (respectively, $g_{int},g_{ext}$) maps are shown with red arrows.
    The left cospan represents a symmetry placed inside a hierarchical edge while the right one depicts a tensor product of two identities put inside the hierarchical edge.
    The colours of the vertices show the induced partition of the interfaces.
    We will try to build an isomorphism step-by-step.
    To make the diagram commute, clearly, it must be the case that $u_0 \mapsto u_0$ and $u_1 \mapsto u_1$.
    Then, because the isomorphism should preserve the order within the subset of green verices, it must be the case that $u_2 \mapsto u_2$ and $u_3 \mapsto u_3$.
    Similarly for the output interfaces.
    To further make the diagram commute, we need to map $v_2 \mapsto v_3$ and $v_3 \mapsto v_2$.
    This, however, contradicts the requirement that $g_{int};\gamma(w_2) = g_{int}(w_2)$.
    Hence, the two cospans are not isomorphic.
\end{example}

### 2025-01-21 LICS/sections/combinatorial_semantics.tex
wip

In ~\cite{bonchi_string_2022-1},  hypergraphs \textit{with interfaces} and their DPO rewriting are shown to be sound and complete for categorical presentations of SMTs including a Frobenius algebra. 
Hypergraphs generalise graphs by allowing edges to have multiple sources and targets. 
When interpreting string diagrams as hypergraphs,  morphisms are interpreted as edges,  and wires as vertices.  
However,  there is also a need to specify which vertices are to be considered as \textit{input} and \textit{outputs},  corresponding to the dangling wires at the bottom and top of a string diagram. 
This gives rise to the definition of a hypergraph with interfaces.  
Subsequent work \cite{bonchi_string_2022-2} builds on this correspondence to give a restriction of hypergraphs with interfaces (called \textit{monogamous} directed acyclic),  leading to a correspondence between symmetric monoidal string diagrams (\textit{i.e.}, without requiring a Frobenius algebra) and the category of such hypergraphs.

@@

Then, morphisms of $\catname{S}(\Sigma)$ can be interpreted as particular cospans of discrete hypergraphs,  which we call hypergraphs with interfaces~\cite{bonchi_string_2022-2}.
\begin{definition}[Hypergraphs with interfaces]
\label{def:cspd}
Define the category of \emph{hypergraphs with interfaces} $\catname{HypI(\Sigma)}$ as the full sub-category of $\catname{Csp(Hyp(\Sigma))}$ with discrete hypergraphs as objects.
\end{definition}

In particular, this means that any morphism in $\HypI{\Sigma}$ is of the form $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$,
where $n,\;m$ are discrete hypergraphs and $\mathcal{G}$ is an arbitrary hypergraph.  The morphisms $f$ and $g$ into $\mathcal{G}$ will specify which vertices are to be considered as inputs and outputs,  respectively.

@@

In order to represent only morphisms of $\catname{S}(\Sigma)$,  it is necessary to restrict our attention to the following subset of hypergraphs with interfaces. 
\begin{definition}[Category of MDA Hypergraphs with Interfaces]
\label{def:monogamy_hyp}
We call a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\HypI{\Sigma}$ \emph{monogamous directed acyclic (MDA)} if:
\begin{enumerate}
    \item $\mathcal{G}$ is a directed acyclic hypergraph;
    \item $f$ and $g$ are monomorphisms;
    \item the in-degree and out-degree of every vertex is at most $1$;
    \item vertices with in-degree $0$ are precisely the image of $f$; and
    \item vertices with out-degree $0$ are precisely the image of $g$.
\end{enumerate}
We denote by $\MdaCospans$ the subcategory of $\HypI{\Sigma}$ of MDA cospans.
\end{definition}
Note that $\MdaCospans$ remains a symmetric monoidal category with the same equipment as before.  We now recall from
\cite{bonchi_string_2022-2} the following theorem,  which says the category $\MdaCospans$ is in fact \textit{equivalent} to $\catname{S}(\Sigma)$. 
\begin{theorem}[Corollary 26 of~\cite{bonchi_string_2022-2}]\label{thm:prop-equiv}
    $\catname{S}(\Sigma) \cong \MdaCospans$
\end{theorem}

@@

\begin{definition}[Boundary complement]
\label{def:boundary_original}
For MDA cospans $i \xrightarrow{a_1} L \xleftarrow{a_2} j$ and $n \xrightarrow{b_1} \mathcal G \xleftarrow{b_2} m$
and mono $f : \mathcal L \to \mathcal G$, a pushout complement $i + j \to \mathcal{L}^\bot \to \mathcal G$ as depicted in the square below
% \begin{figure}[h!]
%     \centering
%     \includegraphics[width=0.4\linewidth]{figures/combinatorial_semantics/boundary_complement.png}
% \end{figure}
% https://q.uiver.app/#q=WzAsNSxbMCwwLCJMIl0sWzIsMCwiaStqIl0sWzAsMiwiR157XFx1cmNvcm5lcn0iXSxbMiw0LCJuK20iXSxbMiwyLCJMXntcXGJvdH0iXSxbMyw0LCJbZF8xLGRfMl0iLDJdLFszLDIsIltiXzEsYl8yXSJdLFswLDIsImYiLDJdLFsxLDAsImEgPSBbYV8xLGFfMl0iLDJdLFsxLDQsImMgPSBbY18xLGNfMl0iXSxbNCwyLCJnIl1d
% \[\begin{tikzcd}
% 	\mathcal L && {i+j} \\
% 	\\
% 	{\mathcal G^{\urcorner}} && {\mathcal{L}^{\bot}} \\
% 	\\
% 	&& {n+m}
% 	\arrow["{[d_1,d_2]}"', from=5-3, to=3-3]
% 	\arrow["{[b_1,b_2]}", from=5-3, to=3-1]
% 	\arrow["f"', from=1-1, to=3-1]
% 	\arrow["{[a_1,a_2]}"', from=1-3, to=1-1]
% 	\arrow["{[c_1,c_2]}", from=1-3, to=3-3]
% 	\arrow[from=3-3, to=3-1]
% \end{tikzcd}\]
\[
    \scalebox{0.8}{
        \tikzfig{../figures/combinatorial_semantics/DPOI_boundary_complement}
    }
\]
is called a boundary complement if $[c_1, c_2]$ is mono and there exist $d_1 : n \to \mathcal{L}^{\bot}$ and $d_2 : m \to \mathcal{L}^{\bot}$ making the above triangle commute and such that
\[
    n + j \xrightarrow{[d_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[d_2,c_1]} m + i
\]
is an MDA cospan. 
\end{definition}
%\begin{remark}
%    Note how the requirement for $i \xrightarrow{} L \xleftarrow{} j$ and $n \xrightarrow{} G \xleftarrow{} m$ being monogamous, i.e. morphisms in $\MdaCospans$, is a part of the above definition.
%\end{remark}
Intuitively,  requiring boundary complements ensures that inputs are glued exclusively to outputs and vice-versa in the two pushout squares.
Yet, this restriction is not enough to make DPOI rewriting complete for $\catname{S,\mathcal{E}}$: there might be cospans with carriers $\mathcal{G}$ and $\mathcal{H}$ such that one can be rewritten into another but their corresponding morphisms are not equal modulo $\mathcal{E}$~\cite{bonchi_string_2022-1}.
To make the DPOI rewriting complete, we need to add a restriction on the image of the match.

@@

\begin{definition}[Convex DPOI rewriting]
\label{def:convex_dpo}
Let $\mathfrak{R}$ be a set of DPOI rewrite rules. 
Then, given $\mathcal G \xleftarrow{} n+m$ and $\mathcal H \xleftarrow{} n + m$ in $\catname{Hyp(\Sigma)}$, $\mathcal G$ rewrites convexly into $\mathcal H$ with interface $n + m$ --- notation $(\mathcal G \xleftarrow{} n + m ) \Rrightarrow_{\mathfrak{R}}  (\mathcal H \xleftarrow{} n + m )$ --- if there exist rule $\mathcal L \xleftarrow{} i + j \xrightarrow{} \mathcal R$ in $\mathfrak{R}$ and object $\mathcal{L}^{\bot}$ and cospan arrows $i+j \xrightarrow{} \mathcal{L}^{\bot} \xleftarrow{} n+m$ such that the DPOI diagram of Definition \ref{def:dpoi} commutes and its marked squares are pushouts 
and the following conditions hold
\begin{itemize}
    \item $f : L \to \mathcal G$ is a convex match;
    \item $i + j \to \mathcal{L}^{\bot} \to \mathcal G$ is a boundary complement in the leftmost pushout.
\end{itemize}
\end{definition}
%By requiring convex matching and boundary complements we ensure that the original cospans above are monogamous, e.g., $n \xrightarrow{} G \xleftarrow{} m$.
We do not formally recite the main result of \cite{bonchi_string_2022-2} here,  but the work proves that for any morphisms $f$ and $g$ in $\catname{S}(\Sigma, \mathcal{E})$,  we have that $f = g$ if and only if there exist a sequence of DPO rewrites (induced by $\mathcal{E}$) between the representations of $f$ and $g$ as hypergraphs.  In particular,  Theorem \ref{thm:prop-equiv} means that the structural SMC equations are factored out in the representation of $f$ and $g$,  and the DPO rewrites necessary are only those induced by $\mathcal{E}$. 
%\begin{theorem}[Theorem 35~\cite{bonchi_string_2022-2}]
%For any $f,g \in \mathbf{SMT}(\Sigma, \mathcal{E})$,  we have that 
%\[
%	f = g \iff \llangle \lceil f \rceil \rrangle \Rrightarrow^*_{\mathfrak{R}} \llangle \lceil g \rceil \rrangle~, 
%\]
%where $\mathfrak{R} = \llangle \lceil \mathcal{E} \rceil \rrangle$ and $ \Rrightarrow^*_{\mathfrak{R}}$ denotes the existence of a sequence of convex rewrites. 
%\end{theorem}

>>>>

In ~\cite{bonchi_string_2022-1},  hypergraphs \textit{with interfaces} and their DPO rewriting are shown to be sound and complete for categorical presentations of SMTs including a Frobenius algebra. 
Hypergraphs generalise graphs by allowing edges to have multiple sources and targets. 
When interpreting string diagrams as hypergraphs,  morphisms are interpreted as edges,  and wires as vertices.  
However,  there is also a need to specify which vertices are to be considered as \textit{input} and \textit{outputs},  corresponding to the dangling wires at the bottom and top of a string diagram. 
This gives rise to the definition of a hypergraph with interfaces.  
Subsequent work \cite{bonchi_string_2022-2} builds on this correspondence to give a restriction of hypergraphs with interfaces (called \textit{monogamous} directed acyclic),  leading to a correspondence between symmetric monoidal string diagrams (\textit{i.e.}, without requiring a Frobenius algebra) and the category of (cospans of) such hypergraphs.

@@

Then, morphisms of $\catname{S}(\Sigma)$ can be interpreted as particular cospans of discrete hypergraphs,  which we call hypergraphs with interfaces~\cite{bonchi_string_2022-2}.
\begin{definition}[Hypergraphs with interfaces]
\label{def:cspd}
Define the category of \emph{hypergraphs with interfaces} $\catname{HypI(\Sigma)}$ as the full sub-category of $\catname{Csp(Hyp(\Sigma))}$ with discrete \textit{ordered} hypergraphs as objects.
\end{definition}

In particular, this means that any morphism in $\HypI{\Sigma}$ is of the form $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$,
where $n,\;m$ are discrete hypergraphs and $\mathcal{G}$ is an arbitrary hypergraph.  The morphisms $f$ and $g$ into $\mathcal{G}$ will specify which vertices are to be considered as inputs and outputs,  respectively.
The ordering of the interface hypergraphs is the way to differentiate between the interpretations of morphisms $f \otimes g$ and $g \otimes f$.

@@

In order to make the interpretation complete with respect to $\catname{S}(\Sigma)$,  it is necessary to restrict our attention to the following subset of hypergraphs with interfaces. 
\begin{definition}[Category of MDA Hypergraphs with Interfaces~\cite{bonchi_string_2022-2}]
\label{def:monogamy_hyp}
We call a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\HypI{\Sigma}$ \emph{monogamous directed acyclic (MDA)} if:
\begin{enumerate}
    \item $\mathcal{G}$ is a directed acyclic hypergraph;
    \item $f$ and $g$ are monomorphisms;
    \item the in-degree and out-degree of every vertex is at most $1$;
    \item vertices with in-degree $0$ are precisely the image of $f$; and
    \item vertices with out-degree $0$ are precisely the image of $g$.
\end{enumerate}
We denote by $\MdaCospans$ the subcategory of $\HypI{\Sigma}$ of MDA cospans.
\end{definition}
Note that $\MdaCospans$ remains a symmetric monoidal category with the same equipment as before.  We now recall from
\cite{bonchi_string_2022-2} the following theorem,  which says the category $\MdaCospans$ is in fact \textit{equivalent} to $\catname{S}(\Sigma)$. 
\begin{theorem}[Corollary 26 of~\cite{bonchi_string_2022-2}]\label{thm:prop-equiv}
    $\catname{S}(\Sigma) \cong \MdaCospans$
\end{theorem}

@@

\begin{definition}[Boundary complement]
\label{def:boundary_original}
For MDA cospans $i \xrightarrow{a_1} L \xleftarrow{a_2} j$ and $n \xrightarrow{b_1} \mathcal G \xleftarrow{b_2} m$
and mono $f : \mathcal L \to \mathcal G$, a pushout complement $i + j \to \mathcal{L}^\bot \to \mathcal G$ as depicted in the square below
% \begin{figure}[h!]
%     \centering
%     \includegraphics[width=0.4\linewidth]{figures/combinatorial_semantics/boundary_complement.png}
% \end{figure}
% https://q.uiver.app/#q=WzAsNSxbMCwwLCJMIl0sWzIsMCwiaStqIl0sWzAsMiwiR157XFx1cmNvcm5lcn0iXSxbMiw0LCJuK20iXSxbMiwyLCJMXntcXGJvdH0iXSxbMyw0LCJbZF8xLGRfMl0iLDJdLFszLDIsIltiXzEsYl8yXSJdLFswLDIsImYiLDJdLFsxLDAsImEgPSBbYV8xLGFfMl0iLDJdLFsxLDQsImMgPSBbY18xLGNfMl0iXSxbNCwyLCJnIl1d
% \[\begin{tikzcd}
% 	\mathcal L && {i+j} \\
% 	\\
% 	{\mathcal G^{\urcorner}} && {\mathcal{L}^{\bot}} \\
% 	\\
% 	&& {n+m}
% 	\arrow["{[d_1,d_2]}"', from=5-3, to=3-3]
% 	\arrow["{[b_1,b_2]}", from=5-3, to=3-1]
% 	\arrow["f"', from=1-1, to=3-1]
% 	\arrow["{[a_1,a_2]}"', from=1-3, to=1-1]
% 	\arrow["{[c_1,c_2]}", from=1-3, to=3-3]
% 	\arrow[from=3-3, to=3-1]
% \end{tikzcd}\]
\[
    \scalebox{0.8}{
        \tikzfig{../figures/combinatorial_semantics/DPOI_boundary_complement}
    }
\]
is called a boundary complement if $[c_1, c_2]$ is mono and there exist $d_1 : n \to \mathcal{L}^{\bot}$ and $d_2 : m \to \mathcal{L}^{\bot}$ making the above triangle commute and such that
\[
    n + j \xrightarrow{[d_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[d_2,c_1]} m + i
\]
is an MDA cospan. 
\end{definition}
%\begin{remark}
%    Note how the requirement for $i \xrightarrow{} L \xleftarrow{} j$ and $n \xrightarrow{} G \xleftarrow{} m$ being monogamous, i.e. morphisms in $\MdaCospans$, is a part of the above definition.
%\end{remark}
Intuitively,  requiring boundary complements ensures that inputs are glued exclusively to outputs and vice-versa in the two pushout squares.
Yet, this restriction is not enough to make DPOI rewriting complete for $\catname{S}(\Sigma,\mathcal{E})$: there might be cospans with carriers $\mathcal{G}$ and $\mathcal{H}$ such that one can be rewritten into another but their corresponding morphisms are not equal modulo $\mathcal{E}$~\cite{bonchi_string_2022-2}.
To make the DPOI rewriting complete, we need to add a restriction on the image of the match.

@@

\begin{definition}[Convex DPOI rewriting]
\label{def:convex_dpo}
Let $\mathfrak{R}$ be a set of DPOI rewrite rules. 
Then, given $\mathcal G \xleftarrow{} n+m$ and $\mathcal H \xleftarrow{} n + m$ in $\catname{Hyp(\Sigma)}$, $\mathcal G$ rewrites convexly into $\mathcal H$ with interface $n + m$ --- notation $(\mathcal G \xleftarrow{} n + m ) \Rrightarrow_{\mathfrak{R}}  (\mathcal H \xleftarrow{} n + m )$ --- if there exist rule $\mathcal L \xleftarrow{} i + j \xrightarrow{} \mathcal R$ in $\mathfrak{R}$ and object $\mathcal{L}^{\bot}$ and cospan arrows $i+j \xrightarrow{} \mathcal{L}^{\bot} \xleftarrow{} n+m$ such that the DPOI diagram of Definition \ref{def:dpoi} commutes and its marked squares are pushouts 
and the following conditions hold
\begin{itemize}
    \item $f : L \to \mathcal G$ is a convex match;
    \item $i + j \to \mathcal{L}^{\bot} \to \mathcal G$ is a boundary complement in the leftmost pushout.
\end{itemize}
\end{definition}
%By requiring convex matching and boundary complements we ensure that the original cospans above are monogamous, e.g., $n \xrightarrow{} G \xleftarrow{} m$.
We do not formally recite the main result of \cite{bonchi_string_2022-2} here,  but the work proves that for any morphisms $f$ and $g$ in $\catname{S}(\Sigma, \mathcal{E})$,  we have that $f = g$ if and only if there exist a sequence of DPO rewrites (induced by $\mathcal{E}$) between the representations of $f$ and $g$ as (cospans of) hypergraphs.
In particular,  Theorem \ref{thm:prop-equiv} means that the structural SMC equations are factored out in the representation of $f$ and $g$,  and the DPO rewrites necessary are only those induced by $\mathcal{E}$. 
%\begin{theorem}[Theorem 35~\cite{bonchi_string_2022-2}]
%For any $f,g \in \mathbf{SMT}(\Sigma, \mathcal{E})$,  we have that 
%\[
%	f = g \iff \llangle \lceil f \rceil \rrangle \Rrightarrow^*_{\mathfrak{R}} \llangle \lceil g \rceil \rrangle~, 
%\]
%where $\mathfrak{R} = \llangle \lceil \mathcal{E} \rceil \rrangle$ and $ \Rrightarrow^*_{\mathfrak{R}}$ denotes the existence of a sequence of convex rewrites. 
%\end{theorem}
